---
title: "R Notebook"
output: html_notebook
---

# Libraries need
```{r}
library(tidyverse)  # data manipulation
library(cluster)    # clustering algorithms
library(factoextra) # clustering algorithms & visualization
library(data.table)
library(ggplot2)
library(lubridate)
library(knitr)
library(leaflet)
library(caret)
library(flexclust)
library(corrplot)
library(geosphere)
library(NbClust)
library(dbplyr)
```

```{r}
getwd()
```

```{r}
# reading the test and train data
trainData <- read.csv("train_data.csv", header = TRUE, sep = ",")
summary(trainData)
cat("\n")
testData <- read.csv("test_data.csv", header = TRUE, sep = ",")
summary(testData)
print('Done loading data')
```


```{r}
# Checking if nan values are present in the dataset
any(is.na(trainData))
```

```{r}
# used for omitting the nan values from the data
trainData <- na.omit(trainData)
```

```{r}
table(trainData['passenger_count'])
```

```{r}
# remove the passenger_count=0
trainData = trainData[trainData$passenger_count != 0, ]
```

```{r}
table(trainData['passenger_count'])
```

```{r}
ggplot(trainData, aes(fare_amount))+
  geom_histogram(fill = "blue", bins = 50)+
  ggtitle("Distribution of Fare Amount")+
  theme(plot.title = element_text(hjust = .5),)
```

# Adding features
```{r}
# function to get distance between points
distance = function(lat1, long1, lat2, long2) {
    deglen = 110.25
    x = lat1 - lat2
    y = (long1 - long2) * cos(lat1)
    return(deglen * sqrt(x * x + y * y))
}
```



```{r}
print("Add distance to trainData data")
trainData$distance = distance(trainData$pickup_latitude, trainData$pickup_longitude, 
    trainData$dropoff_latitude, trainData$dropoff_longitude)
cat("\n")
print("Add distance to testData data")
testData$distance = distance(testData$pickup_latitude, testData$pickup_longitude, 
    testData$dropoff_latitude, testData$dropoff_longitude)

```


```{r}
# Add hour of pickup
print("Add hour feature to trainData data")
# To train data
t.lub.train <- ymd_hms(trainData$pickup_datetime)
trainData$pickup_hour <- as.numeric(format(t.lub.train, "%H")) + as.numeric(format(t.lub.train, "%M"))/60

cat("\n")
print("Add hour feature to testData data")
# To test data
t.lub.test <- ymd_hms(testData$pickup_datetime)
testData$pickup_hour <- as.numeric(format(t.lub.test, "%H")) + as.numeric(format(t.lub.test, "%M"))/60
```

```{r}
# Add month
print("Add month feature to test and train data")
trainData$month <- as.factor(format(t.lub.train, "%m"))
testData$month <- as.factor(format(t.lub.test, "%m"))
```


```{r}
head(trainData)
```

# Checking for outliers

```{r}
hist(trainData$distance,main = 'Distance (km)',breaks=c(0,5,10,15,20,25,30,35,500000),xlim=c(0,40))
```

```{r}
boxplot(trainData$distance,main = 'Distance (km)',ylim=c(0,20))
```


# From the graph, removing the outliers any items with distance > 20 km, and also distance of 0 km.
```{r}
trainData = trainData[trainData$distance != 0, ]

trainData = trainData[trainData$distance < 20, ]
```


# Leaflet Sample Display

```{r}
# A map view of 1000 random Pickup locations using leaflet is shown below:
sample_data = trainData[sample(1:dim(trainData)[1],1000),]

leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addAwesomeMarkers(
    lng=sample_data$pickup_longitude, 
    lat=sample_data$pickup_latitude,
    clusterOptions = markerClusterOptions()
    
  )
```


```{r}
# Map view of corresponding drop off locations using leaflet
leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addAwesomeMarkers(
    lng=sample_data$dropoff_longitude, 
    lat=sample_data$dropoff_latitude,
    clusterOptions = markerClusterOptions()
  )
```


# Clustering

```{r}
# Define number of clusters
cen = 10

pickup = rbind(cbind(trainData$pickup_longitude,trainData$pickup_latitude),cbind(testData$pickup_longitude,testData$pickup_latitude))
clus_pickup = kmeans(pickup, centers = cen, nstart = 1)
```


```{r}
# adding feature pickupcluster and dropoffcluster to the dataset
trainData$pickupclus = as.factor(clus_pickup$cluster[1:dim(trainData)[1]])
testData$pickupclus = as.factor(clus_pickup$cluster[(dim(trainData)[1]+1):length(clus_pickup$cluster)])


# cluster locations for dropoff

dropoff = rbind(cbind(trainData$dropoff_longitude,trainData$dropoff_latitude),cbind(testData$dropoff_longitude,testData$dropoff_latitude))
clus_dropoff = kmeans(dropoff, centers = cen, nstart = 1)

trainData$dropoffclus = as.factor(clus_dropoff$cluster[1:dim(trainData)[1]])
testData$dropoffclus = as.factor(clus_dropoff$cluster[(dim(trainData)[1]+1):length(clus_dropoff$cluster)])
```


```{r}
sample_data = trainData[sample(1:dim(trainData)[1],100000),]

ggplot(mapping=aes(x=sample_data$pickup_longitude,y=sample_data$pickup_latitude))+
  geom_point(alpha=.2,cex=0.0001,aes(colour=sample_data$pickupclus))+
  scale_x_continuous(limits = c(-74.02,-73.85))+
  scale_y_continuous(limits = c(40.7,40.85))+
  ggtitle('Pick up locations clustered')+
  guides(fill=F)+
  xlab('Longitude')+ylab('Latitude')+ggtitle('Plot of pick up locations with clustering')+
  theme(legend.position = 'none')
```



```{r}
# dropoff location by cluster----
  ggplot(mapping=aes(x=sample_data$dropoff_longitude,y=sample_data$dropoff_latitude))+
    geom_point(alpha=.2,cex=0.0001,aes(colour=sample_data$dropoffclus))+
    scale_x_continuous(limits = c(-74.02,-73.85))+
    scale_y_continuous(limits = c(40.7,40.85))+
  ggtitle('Drop off locations clustered')+
  theme(legend.position = 'none')+
  xlab('Longitude')+ylab('Latitude')+ggtitle('Plot of drop off locations with clustering')
```


```{r}
summary(trainData)
```


# Correlation plot

# getting subset of the dataframe
```{r}
df <- trainData[,c("fare_amount","pickup_longitude","pickup_latitude", "dropoff_longitude","dropoff_latitude","passenger_count", "distance")] 
```


```{r}
# TO plotcorrelations between predictors
predictorCorr = cor(df[,-1])
corrplot(predictorCorr, method = "number")
```

# positive or negative correlation which over estimates or under estimates in linear regression

```{r}
# Pearson correlation between 2 variables
cor(df$fare_amount, df$distance)
```


```{r}
# correlation between 2 variables
cor(df$fare_amount, df$passenger_count)
```

```{r}
# correlation for all variables
round(cor(df),
  digits = 2 # rounded to 2 decimals
)
```




# Converting character into datatime
```{r}
testData$pickup_datetime <- ymd_hms(testData$pickup_datetime)
```


```{r}
testData$key <- ymd_hms(testData$key)
```


# getting subset of the dataframe
```{r}
testdf <- testData[,c("pickup_longitude","pickup_latitude", "dropoff_longitude","dropoff_latitude","passenger_count", "distance")] 
print(testdf)
```

```{r}
# k-means with 2 clusters
k2 <- kmeans(df, centers = 2, nstart = 1)
str(k2)
```

```{r}
summary(df)
```


```{r}
k3 <- kmeans(df, centers = 3, nstart = 25)
k4 <- kmeans(df, centers = 4, nstart = 25)
k5 <- kmeans(df, centers = 5, nstart = 25)

# plots to compare
p1 <- fviz_cluster(k2, geom = "point", data = df) + ggtitle("k = 2")
p2 <- fviz_cluster(k3, geom = "point",  data = df) + ggtitle("k = 3")
p3 <- fviz_cluster(k4, geom = "point",  data = df) + ggtitle("k = 4")
p4 <- fviz_cluster(k5, geom = "point",  data = df) + ggtitle("k = 5")

library(gridExtra)
grid.arrange(p1, p2, p3, p4, nrow = 2)
```

```{r}
k6 <- kmeans(df, centers = 6, nstart = 25)
k7 <- kmeans(df, centers = 7, nstart = 25)

p5 <- fviz_cluster(k6, geom = "point",  data = df) + ggtitle("k = 6")
p6 <- fviz_cluster(k7, geom = "point",  data = df) + ggtitle("k = 7")

grid.arrange(p5, p6, nrow = 2)
```


```{r}
# Elbow method
set.seed(123)

# function to compute total within-cluster sum of square 
wss <- function(k) {
  kmeans(df, k, nstart = 10 )$tot.withinss
}

# Compute and plot wss for k = 1 to k = 15
k.values <- 1:15

# extract wss for 2-15 clusters
wss_values <- map_dbl(k.values, wss)

plot(k.values, wss_values,
       type="b", pch = 15, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")
```


# Considering Prime locations from the dataset

```{r}
trainData<-trainData%>%  
  filter(pickup_longitude > -80 & pickup_longitude < -70) %>%
  filter(pickup_latitude > 35 & pickup_latitude < 45) %>%
  filter(dropoff_longitude > -80 & dropoff_longitude < -70) %>%
  filter(dropoff_latitude > 35 & dropoff_latitude < 45)
```

```{r}
traindf1 <- trainData[,c("key","fare_amount","pickup_datetime","pickup_longitude","pickup_latitude", "dropoff_longitude","dropoff_latitude","passenger_count", "distance")] 
print(traindf1)
```



```{r}
testdf1 <- testData[,c("key","pickup_datetime","pickup_longitude","pickup_latitude", "dropoff_longitude","dropoff_latitude","passenger_count", "distance")] 
print(testdf1)
```


```{r}
testData$fare_amount<-NA
key<-testdf1$key
testdf1<-testdf1%>%
  select(-key)
```


```{r}
all<-data.frame(rbind(trainData, testData))


all<-all%>%
  mutate(
    pickup_datetime = ymd_hms(pickup_datetime),
    year = as.factor(year(pickup_datetime)),
    month = as.factor(month(pickup_datetime)),
    day = as.numeric(day(pickup_datetime)),
    dayOfWeek = as.factor(wday(pickup_datetime)),
    hour = as.numeric(hour(pickup_datetime)),
    #minute = as.numeric(minute(pickup_datetime)),
    #quarter = as.factor(quarter(pickup_datetime)),
    timeOfDay = as.factor(ifelse(hour >= 3 & hour < 9,
                    "Morning", ifelse(hour >= 9 & hour < 14, "Mid-Day",
                        ifelse(hour >= 14 & hour < 18, "Evening", "Night"))))
  )%>%
  select(-pickup_datetime)
```


```{r}
#jfk
jfk_lat<-40.6413
jfk_long<--73.7781
jfk<-c(jfk_long, jfk_lat)
#newark
nwk_lat<-40.6895
nwk_long<--74.1745
nwk<-c(nwk_long, nwk_lat)
#laguardia
lag_lat<-40.779
lag_long<--73.8740
lag<-c(lag_long, lag_lat)
#MSG
msg_lat<-40.7505
msg_long<--73.9934
msg<-c(msg_long, msg_lat)
#times square
ts_lat<-40.7589
ts_long<--73.9851
ts<-c(ts_long, ts_lat)
#freedom tower
freedom_lat<-40.7127
freedom_long<--74.0134
freedom<-c(freedom_long, freedom_lat)
#empire state building
esb_lat<-40.7484
esb_long<--73.9857
esb<-c(esb_long, esb_lat)
#grand central
grand_lat<-40.7527
grand_long<--73.9772
grand<-c(grand_long, grand_lat)
#bronx
bronx_lat <- (40.837048 * pi)/180
bronx_long <- (-73.865433 * pi)/180
bronx<-c(bronx_long, bronx_lat)
nyc<-c(-74.0063889, 40.7141667)



all<-all%>%
  mutate(
    dist = distHaversine(cbind(pickup_longitude, pickup_latitude), cbind(dropoff_longitude, dropoff_latitude), r = 6371),
    to_jfk = distHaversine(cbind(pickup_longitude, pickup_latitude), jfk, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), jfk, r = 6371),
    to_nkw = distHaversine(cbind(pickup_longitude, pickup_latitude), nwk, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), nwk, r = 6371),
    to_lag = distHaversine(cbind(pickup_longitude, pickup_latitude), lag, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), lag, r = 6371),
    to_msg = distHaversine(cbind(pickup_longitude, pickup_latitude), msg, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), msg, r = 6371),
    to_ts = distHaversine(cbind(pickup_longitude, pickup_latitude), ts, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), ts, r = 6371),
    to_freedom = distHaversine(cbind(pickup_longitude, pickup_latitude), freedom, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), freedom, r = 6371),
  
    to_grand = distHaversine(cbind(pickup_longitude, pickup_latitude), grand, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), grand, r = 6371),
    to_bronx = distHaversine(cbind(pickup_longitude, pickup_latitude), bronx, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), bronx, r = 6371),
    to_nyc = distHaversine(cbind(pickup_longitude, pickup_latitude), nyc, r = 6371) + distHaversine(cbind(dropoff_longitude, dropoff_latitude), nyc, r = 6371)
    
    
  )
```


```{r}
#Separate train from validation
set.seed(0)
size = floor(.8*nrow(na.omit(all)))

ii<-sample(1:nrow(na.omit(all)), size)
train<-na.omit(all)[ii,]
valid<-na.omit(all)[-ii,]

#train$pickup_datetime <- ymd_hms(train$pickup_datetime)
train$key <- ymd_hms(train$key)


dist_var<-grep("to_", colnames(train))
dist_df<-train[,dist_var]
print(train)
```

```{r}
train <- train[,c("pickup_longitude","pickup_latitude", "dropoff_longitude","dropoff_latitude","passenger_count")]
print(train)
```

```{r}
#correlation with target (spearman)
correlations_target<-cor(train[,1], dist_df, method = "spearman")
correlations_target
```


```{r}
correlations_dist<-cor(dist_df, method = "spearman")
corrplot::corrplot(correlations_dist, type = "upper")
```

```{r}
pcaObject<-prcomp(dist_df, center = TRUE, scale. = TRUE)


fviz_screeplot(pcaObject)
```

```{r}
a1<-fviz_contrib(pcaObject, choice = "var", axes = 1)
b1<-fviz_contrib(pcaObject, choice = "var", axes = 2)
c1<-fviz_contrib(pcaObject, choice = "var", axes = 3)
d1<-fviz_contrib(pcaObject, choice = "var", axes = 4)

a2<-fviz_cos2(pcaObject, choice = "var", axes = 1)
b2<-fviz_cos2(pcaObject, choice = "var", axes = 2)
c2<-fviz_cos2(pcaObject, choice = "var", axes = 3)
d2<-fviz_cos2(pcaObject, choice = "var", axes = 4)

gridExtra::grid.arrange(a1,a2)
```

```{r}
gridExtra::grid.arrange(b1,b2)
```

```{r}
gridExtra::grid.arrange(c1,c2)
```

```{r}
gridExtra::grid.arrange(d1,d2)
```

# getting subset of the dataframe
```{r}
df <- traindf1[,c("fare_amount","pickup_longitude","pickup_latitude", "dropoff_longitude","dropoff_latitude","passenger_count", "distance")] 
```

# Linear model
```{r}
# Linear model
triplinearModel = lm(fare_amount~., data = df)
```


# Calculating the training MSE/RMSE and R-square

```{r}
trainmse = mean((df$fare_amount - triplinearModel$fitted.values)^2)

# Results
cat("Train MSE: ", trainmse)
cat("\n")
cat("Train RMSE: ", sqrt(trainmse))
cat("\n")
cat("Train R-squared: ",summary(triplinearModel)$r.sq)
```



```{r}
# Resulting Fit
plot(x=predict(triplinearModel), y= df$fare_amount,
     xlab='Predicted Values',
     ylab='Actual Values',
     main='Predicted vs. Actual Values')
abline(a=0,b=1, col="red")
```









